Learning how to act: making good decisions with machine learning

### 2017-10-20 thesis/thesis_map.tex
getting ready for talk
finalised talk, replaced most instances of the word framework, fixed spelling and repeated word errors, completes aknowledgements.

%There are many people without whose support this thesis would not have been possible. I would like to thank my supervisor Dr Cheng Song Ong,

% Dr Mark Reid, Dr Tiberio Ceatano and Prof. Robert Williamson.

%My brother Tor for many interesting discussions on bandits.

%To my daughters, Freya and Anouk,

% supervisors
% family (Tor)
% thesis boot camp
% scholarship

%I am grateful to have recieved funding from an Australian Postgraduate Award and NICTA/Data61 scholarship to support this research.

@@

Causality is often linked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisation: the amount of information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distill how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}

@@

\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution is a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.

@@

In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems: a framework that unifies causal graphical models with multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of the bandit problem depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms. In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of the causal bandit problem, which I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.

@@

A central component of estimating causal effects from observational data is determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured. We did this on an ad-hoc basis to resolve the examples discussed in section \ref{sec:causal_models}. In this section we describe a principled approach to mapping observational quantities to interventional ones and then, in section \ref{sec:causal_effect_under_ignorability}, discuss the key issues involved in estimating such expressions from finite sample data. We assume the basic structure of the graph is known. That is, we assume that we can draw a network containing (at a minimum):

@@

\vspace*{.3cm}
\begin{definition}[Non-parametric identifiablity]
Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$. A causal query of the form $\P{\boldsymbol{Y}|do(\boldsymbol{X}),\boldsymbol{W}}$, where $\boldsymbol{Y}$,$\boldsymbol{X}$ and $\boldsymbol{W}$ are disjoint subsets of $\boldsymbol{V}$, is non-parametrically identifiable if it is uniquely determined by $\P{\cdot}$ and $G$.
\end{definition}

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id2}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$X$};
\node[ob, right=of 1](2){$Z$};
\node[ob, right=of 2](3){$Y$};
\node[lt,above=of 2](4){$U$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3)
    (4) edge (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[lt](1){$U$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid2_instrumental_variable}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[lt](1){$U$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[ob, above left=of 2](4){$Z$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (2);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1)
    (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
\label{fig:identify}
\end{figure}

@@

The significance of this should not be underestimated. Cross-validation has allowed applied machine learning to succeed with a very atheoretical approach on the basis that we can identify when a model is successful. With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention. \citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem. There is relatively little theory on model selection for estimating the propensity score. To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion. It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}. With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct. I have found no examples of such an approach.

@@

Undertaking a Bayesian regression of $X$ and $Z$ against $Y$ and using intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ will lead to selection of something centered around $w_{yz}$. However, this is incorrect and can introduce bias into our estimate of the coefficient for $w_{yx}$, as demonstrated in figure \ref{fig:careful-prior-example}. The prior for $\beta$ should be based on the belief about how $Z$ is expected to be \emph{associated} with $Y$, after marginalising out any confounding variables.

\begin{figure}[ht]
\begin{subfigure}[t]{0.99\textwidth}
     \caption{No prior on $\beta$, the posterior on $w_{yx}$ is centered around its true value}
    \includegraphics[scale=1]{prior_no_prior}
\end{subfigure}
\begin{subfigure}[t]{0.99\textwidth}
     \caption{Prior on $\beta$ centered around $w_{yz}$, the causal effect of $Z$ on $Y$, $Prior(\beta)= N(w_{yz},\sigma = 0.5)$. The posterior on $w_{yx}$ is biased away from its true value (as is the posterior on $\beta$ but this is not the key quantity of interest).}
    \includegraphics[scale=1]{prior_bad_prior.pdf}
\end{subfigure}
\caption{An example demonstrating how selecting a prior for a nuisance parameter centered around the causal effect of that parameter on the outcome, rather than its association with the outcome, can bias estimates for actual parameter of interest. We sample $n=1000$ data points from the joint distribution defined by $P(U)P(Z|U)P(X|Z)P(Y|U,Z,X) \sim N(0,1)N(2U,0.09)N(0.5Z,1)N(3U -Z + 0.5X , 0.25)$ and fit the model $Y \sim N(w_yx X+\beta Z,\epsilon)$ in Stan and plot the posterior distributions over $w_yx$ and $\beta$. Figure (a) shows the results with no prior (equivalently an improper prior) for both $w_yx$ and $\beta$. Figure (b) shows the results when we place a Gaussian prior centered around the $w_yz$ on the distribution for $\beta$. All models were fit using Stan \citep{carpenter2016stan}}\label{fig:careful-prior-example}
\end{figure}

@@

Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest. If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds. We can then estimate the average causal effects directly from the differences in outcome across treatments. More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, (figure \ref{fig:random_experiment_network_stratified}). This necessitates that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability (\S\ref{sec:causal_effect_under_ignorability}). The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome). See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.

@@

\vspace{0.5cm}
\begin{definition}[Stochasitic k-armed bandit problem]
Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{\cf{Y}{1},...,\cf{Y}{k}}$ be a joint distribution over the rewards for each action. The multi-armed bandit problem proceeds over $T$ rounds. In each round $t$, 
\begin{enumerate}
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous timesteps and a (potentially stochastic) \emph{policy} $\pi$
\item the world stochastically generates the rewards for each action, $[\cfb{Y}{1},...,\cfb{Y}{k}] \sim \P{\vec{y}}$
\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
\end{enumerate}

@@

\begin{figure}
\includegraphics[scale=1]{upper_bound_vs_actual_average_regret}
\caption{The actual performance of the UCB algorithm can be substantially better than suggested by the upper bound, particularly for small $T$. The solid curve shows the mean expected regret associated with the sequence of arms chosen by UCB-1 with $k=2$ arms and the rewards sampled from $bernoulli([.3,.7])$ over 1000 simulations. The dashed curve shows the corresponding upper bound given by the minimum of $T\Delta_{max}$ and equation\ref{eqn:ucb_problem_dependent_regret}.}
\label{fig:ucb_upperbound_vs_actual_performance}
\end{figure}

@@

\section{A framework to connect bandit problems with observational causal inference}
\label{sec:intro_causal_bandit}
A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph. Each possible assignment of variables to values is a potential action (or bandit arm), see figure \ref{fig:unify_frameworks} for a simple example. In some settings, it makes sense to restrict the action space available to the agent to a subset of all the possible actions, for example the set of single variable interventions. The reward could be a general function of the action selected and the final state of the graph. However, for simplicity, we will consider the reward to be the value of a single specified node. We refer to these problems as \emph{causal bandit problems}. In this thesis, I focus on the case where the causal graph is known. Extending this work to simultaneously learning the casual graph is discussed in \S\ref{sec:causal_bandit_discussion}.

@@

The classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$. The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$. Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward. However, the number of actions or arms grows exponentially with the number of variables in the graph making it important to develop algorithms that leverage the graph structure and additional observations.

@@

Our framework bears a superficial similarity to contextual bandit problems, \S\ref{sec:contextual-bandits}, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention. 
However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.

@@

%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret. They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
% Partial monitoring is a very general framework for for decoupling the feedback from the action and reward. It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \citep{Bartok2014}. Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.

@@

I have developed a framework that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this framework encodes some key existing problems in the literature. I showed that knowledge of the causal structure (but not the functional relationships) between variables can induce a novel form of structure between alternate actions and that an algorithm that leverages this structure obtains better performance that one that does not.

>>>>

There are many people without whose support this thesis would not have been possible. I would like to thank my supervisors Dr Cheng Song Ong, Dr Mark Reid and Dr Tiberio Caetano, Prof. Robert Williamson for chairing my panel, my brother Tor for many interesting and insightful discussions on bandit problems, and my parents for helping me proofread and the many days given to care for my lovely daughters. I would also like to thank Victoria, Inger, Natalie and all the people at ANU thesis bootcamp for giving me the techniques I need to get past the blank page and get words out. Finally I want to thank my wonderful husband for his unwavering support and willingness to take on the household chaos, and my two beautiful daughters Freya and Anouk for their patience while mummy worked on her "tesis".

I am also grateful for the financial support provided by an Australian Postgraduate Award and a NICTA/Data61 top up scholarship.

@@

Causality is often linked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisation: the amount of information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}

@@

\paragraph{Contributions}  The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This synthesis allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.

@@

In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems that unify causal graphical models and multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how this framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of the bandit problem depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms. In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of the causal bandit problem, which I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known a priori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. I also show how this measure can be used to quantify the value of optimised interventional data over purely observational data. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of both the parallel and general causal bandit algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.

@@

A central component of estimating causal effects from observational data is determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured. We did this on an ad hoc basis to resolve the examples discussed in section \ref{sec:causal_models}. In this section we describe a principled approach to mapping observational quantities to interventional ones and then, in section \ref{sec:causal_effect_under_ignorability}, discuss the key issues involved in estimating such expressions from finite sample data. We assume the basic structure of the graph is known. That is, we assume that we can draw a network containing (at a minimum):

@@

\vspace*{.3cm}
\begin{definition}[Non-parametric identifiability]
Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$. A causal query of the form $\P{\boldsymbol{Y}|do(\boldsymbol{X}),\boldsymbol{W}}$, where $\boldsymbol{Y}$,$\boldsymbol{X}$ and $\boldsymbol{W}$ are disjoint subsets of $\boldsymbol{V}$, is non-parametrically identifiable if it is uniquely determined by $\P{\cdot}$ and $G$.
\end{definition}

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id2}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$X$};
\node[ob, right=of 1](2){$Z$};
\node[ob, right=of 2](3){$Y$};
\node[lt,above=of 2](4){$U$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3)
    (4) edge (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[lt](1){$U$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid2_instrumental_variable}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[lt](1){$U$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[ob, above left=of 2](4){$Z$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (2);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1)
    (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\caption{Examples of identifiable and non-identifiable queries. In sub figures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In sub figures (d), (e) and (f) it is not.}
\label{fig:identify}
\end{figure}

@@

The significance of this should not be underestimated. Cross-validation has allowed applied machine learning to succeed with a very a theoretical approach on the basis that we can identify when a model is successful. With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention. \citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem. There is relatively little theory on model selection for estimating the propensity score. To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion. It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}. With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct. I have found no examples of such an approach.

@@

Undertaking a Bayesian regression of $X$ and $Z$ against $Y$ and using intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ will lead to selection of something centred around $w_{yz}$. However, this is incorrect and can introduce bias into our estimate of the coefficient for $w_{yx}$, as demonstrated in figure \ref{fig:careful-prior-example}. The prior for $\beta$ should be based on the belief about how $Z$ is expected to be \emph{associated} with $Y$, after marginalising out any confounding variables.

\begin{figure}[ht]
\begin{subfigure}[t]{0.99\textwidth}
     \caption{No prior on $\beta$, the posterior on $w_{yx}$ is centred around its true value}
    \includegraphics[scale=1]{prior_no_prior}
\end{subfigure}
\begin{subfigure}[t]{0.99\textwidth}
     \caption{Prior on $\beta$ centred around $w_{yz}$, the causal effect of $Z$ on $Y$, $Prior(\beta)= N(w_{yz},\sigma = 0.5)$. The posterior on $w_{yx}$ is biased away from its true value (as is the posterior on $\beta$ but this is not the key quantity of interest).}
    \includegraphics[scale=1]{prior_bad_prior.pdf}
\end{subfigure}
\caption{An example demonstrating how selecting a prior for a nuisance parameter centred around the causal effect of that parameter on the outcome, rather than its association with the outcome, can bias estimates for actual parameter of interest. We sample $n=1000$ data points from the joint distribution defined by $P(U)P(Z|U)P(X|Z)P(Y|U,Z,X) \sim N(0,1)N(2U,0.09)N(0.5Z,1)N(3U -Z + 0.5X , 0.25)$ and fit the model $Y \sim N(w_yx X+\beta Z,\epsilon)$ in Stan and plot the posterior distributions over $w_yx$ and $\beta$. Figure (a) shows the results with no prior (equivalently an improper prior) for both $w_yx$ and $\beta$. Figure (b) shows the results when we place a Gaussian prior centred around the $w_yz$ on the distribution for $\beta$. All models were fit using Stan \citep{carpenter2016stan}}\label{fig:careful-prior-example}
\end{figure}

@@

Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest. If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds. We can then estimate the average causal effects directly from the differences in outcome across treatments. More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, (figure \ref{fig:random_experiment_network_stratified}). This necessitates that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under ignorability (\S\ref{sec:causal_effect_under_ignorability}). The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome). See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.

@@

\vspace{0.5cm}
\begin{definition}[Stochastic k-armed bandit problem]
Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{\cf{Y}{1},...,\cf{Y}{k}}$ be a joint distribution over the rewards for each action. The multi-armed bandit problem proceeds over $T$ rounds. In each round $t$, 
\begin{enumerate}
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous timesteps and a (potentially stochastic) \emph{policy} $\pi$
\item the world stochastically generates the rewards for each action, $[\cfb{Y}{1},...,\cfb{Y}{k}] \sim \P{\vec{y}}$
\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
\end{enumerate}

@@

\begin{figure}
\includegraphics[scale=1]{upper_bound_vs_actual_average_regret}
\caption{The actual performance of the UCB algorithm can be substantially better than suggested by the upper bound, particularly for small $T$. The solid curve shows the mean expected regret associated with the sequence of arms chosen by UCB-1 with $k=2$ arms and the rewards sampled from $Bernoulli([.3,.7])$ over 1000 simulations. The dashed curve shows the corresponding upper bound given by the minimum of $T\Delta_{max}$ and equation\ref{eqn:ucb_problem_dependent_regret}.}
\label{fig:ucb_upperbound_vs_actual_performance}
\end{figure}

@@

\section{A synthesis that connects bandit problems and observational causal inference}
\label{sec:intro_causal_bandit}
A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph. Each possible assignment of variables to values is a potential action (or bandit arm), see figure \ref{fig:unify_frameworks} for a simple example. In some settings, it makes sense to restrict the action space available to the agent to a subset of all the possible actions, for example the set of single variable interventions. The reward could be a general function of the action selected and the final state of the graph. However, for simplicity, we will consider the reward to be the value of a single specified node. We refer to these problems as \emph{causal bandit problems}. In this thesis, I focus on the case where the causal graph is known. Extending this work to simultaneously learning the casual graph is discussed in \S\ref{sec:causal_bandit_discussion}.

@@

The classical $K$-armed stochastic bandit problem can be recovered in the causal bandit setting by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$. The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$. Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward. However, the number of actions or arms grows exponentially with the number of variables in the graph making it important to develop algorithms that leverage the graph structure and additional observations.

@@

Causal bandit problems have a superficial similarity to contextual bandit problems, \S\ref{sec:contextual-bandits}, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention. 
However, a crucial difference is that we have focused on analysing settings where we receive post-action feedback, in which the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.

@@

%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimise cumulative regret. They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
 Partial monitoring is a very general framework for decoupling the feedback from the action and reward. It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \citep{Bartok2014}. Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as the number of actions, which is the key focus of incorporating causal structure.

@@

I have developed an approach that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this conceptualisation encodes some key existing problems in the literature. I showed that knowledge of the causal structure (but not the functional relationships) between variables can induce a novel form of structure between alternate actions and that an algorithm that leverages this structure obtains better performance that one that does not. I introduce a metric that captures the difficulty of causal bandit problems and (under strong assumptions) allows the value of optimised interventional data over purely observational data to be computed prior to intervening in the system.

This work represents an important first step towards a unified approach to causal inference and optimal decision making. There is much exciting work remaining to be done. Although the causal bandit setting can capture contextual information as well as post-action feedback, I have formally analysed and developed algorithms only for the latter. Additionally, to make the problem tractable, I made the (major) assumption that the interventional distribution over the parents of the outcome was known. This can be relaxed to assuming the interventional distribution over some Markov blanket with respect to the outcome is known. Information can then be shared between actions outside the blanket in the same way as in algorithm \ref{alg:general}, whilst actions inside could be learned explicitly. Relaxing this assumption entirely is a much more challenging problem. However, as is demonstrated by the specific example of the parallel causal bandit problem, it is possible to develop algorithms that require only the causal structure of the graph and yield substantially lower regret.
